\mythesistitleNOTSYNPOSIS

### 2012-02-28 src/030chap3.tex
finished proof reading. added ack.

\begin{definition}[Path labeling of a hypergraph]
  \label{def:pathlabel}
  Consider a hypergraph $\cF$ % with universe $U$
  and a path system $\cP$ from tree $T$ such that $\bI(\cF) \cong
  \bI(\cP)$.  Then the associated bijection $\cl: \cF
  \rightarrow \cP$ due to this isomorphism is called a {\em path
    labeling} of $\cF$ from $T$. The path system $\cP$ is
  alternatively denoted as $\cF^\cl$. The path labeling may also be
  denoted by ordered pairs $(\cF,\cl)$ or $(\cF,\cl,T)$.
\end{definition}
% If the intersection graphs of $\cF$ and $\cP$ are isomorphic,
% $\bI(\cF) \cong \bI(\cP)$, then the associated bijection $\cl: \cF
% \rightarrow \cP$ due to this isomorphism is called a {\em path
%   labeling} of the hypergraph $\cF$.
To elaborate on the phrase ``associated bijection'' in
Definition~\ref{def:pathlabel}, let $\cg: V(\bI(\cF)) \rightarrow
V(\bI(\cP))$ be the given intersection graph isomorphism, where
$V(\bI(\cF)) = \{ v_S \mid S \in \cF\}$ and $V(\bI(\cP)) = \{v_P \mid
P \in \cP\}$. Then the associated bijection $\cl: \cF \rightarrow \cP$
is defined as follows. For all $S' \in \cF$ and $P' \in \cP$, $\cl(S')
= P'$ \iff $\cg(v_{S'}) = v_{P'}$.  Thus, it can be seen that a path
labeling preserves the non-empty intersection property between every
pair of hyperedges in $\cF$.

@@

\begin{definition}[Hypergraph isomorphism induced path labeling]
  \label{def:inducedpl}
  Let $\cF$ be a hypergraph and $\cP$ be a path system \stt $\cF \cong
  \cP$. Let  $\phi: \supt{\cF} \rightarrow \supt{\cP}$ be this hypergraph
  isomorphism function. % described in
  % Definition~\ref{def:pathhypergraph}.
  Then there is an {\em induced path labeling} % denoted by $\cl_\phi:
%   \cF \rightarrow \cP$ 
  denoted by $\cp_\phi: \cF \rightarrow \cP$
  of the hypergraph \stt % $\cl_\phi(S) = \phi(S)$
  $\cp_\phi(S) = \phi(S)$
  (see Definition~\ref{def:hisomorph}) for all $S \in \cF$.  
\end{definition}

@@

\begin{definition}[Feasible path labeling]
  \label{def:feasibletpl}
  Let $\cF$ be a hypergraph and $(\cF, \cl)$ be a path
  labeling. $(\cF, \cl)$ is called a {\em feasible path labeling} if
  the following conditions hold.
  \begin{itemize}
  \item $\cF \cong \cF^\cl$
  \item if the above hypergraph isomorphism is $\phi: supp(\cF)
    \rightarrow supp(\cF^\cl)$, then $\cp_\phi = \cl$
  \end{itemize}

@@

Now we will see the definition of a {\em path hypergraph} which is a
simple idea extended from interval hypergraphs in \cite{kklv10}.

\begin{definition}[Path hypergraph]
  \label{def:pathhypergraph}
  Let $\cF$ be a set system. $\cF$ is called a {\em path hypergraph}
  if there exists a target tree $T$ and a path system $\cP$ from $T$
  \stt $\cF \cong \cP$. If $\cF$ is thus a path hypergraph, $\cP$ is
  called {\em path representation} of $\cF$.
\end{definition}

@@

\begin{definition}[$k$-subdivided star]
  \label{def:ksubstar}
  A {\em $k$-subdivided star} is a star with all its rays subdivided
  exactly $k$ times. A {\em ray} of a $k$-subdivided star is the path
  from the center vertex to a leaf. It is clear that all rays of a
  $k$-subdivided star are of length $k+2$.
\end{definition}

@@

\begin{definition}[In-tree]
An {\em in-tree} is a directed rooted tree in which all edges are
directed toward to the root.  
\end{definition}
In-trees are used in Section~\ref{sec:norestraint} which describes our
(albeit non-polynomial time) solution to \CFTPL problem.

@@

\begin{definition}
  \label{def:icppl}
  Consider a path labeling $(\cF, \cl)$ on the given tree $T$. We call
  $(\cF, \cl)$ an {\em Intersection Cardinality Preserving Path
    Labeling (ICPPL)} if it has the following properties.
  \par
  \parbox{\linewidth}{%
    \hangindent \defindent%
    \begin{enumerate}[i. ] %{(\icpplpr\ }
      \singlespacing
    \item \label{pr:i} $|S| = |\cl(S)|$ for all $S \in \cF$
    \item \label{pr:ii}$|S_1 \cap S_2| = |\cl(S_1) \cap \cl(S_2)|$ for
    all distinct $S_1, S_2 \in \cF$
    \item \label{pr:iii}$|S_1 \cap S_2 \cap S_3| = |\cl(S_1) \cap
    \cl(S_2) \cap \cl(S_3)|$ for all distinct $S_1, S_2, S_3 \in \cF$
    \end{enumerate}
  }
\end{definition}

@@

Finally we have Theorem~\ref{th:charac} which puts together all that
we saw so far in this section to prove that ICPPL is indeed a
characterization of a feasible path labeling.
\begin{theorem}
  \label{th:charac}
  A path labeling $(\cF, \cl)$ on tree $T$ is feasible \iff it is an
  ICPPL and Algorithm~\ref{al:icppl-find-isomorph} with $(\cF, \cl,
  T)$ as input returns a non-empty function.
\end{theorem}
\begin{proof}\thesisspacing
  From Lemma~\ref{lem:hyperiso}, we know that if $(\cF, \cl)$ is an
  ICPPL and if Algorithm~\ref{al:icppl-find-isomorph} with $(\cF, \cl,
  T)$ as input returns a non-empty function, then $(\cF, \cl)$ is
  feasible.  Now consider the case where $(\cF, \cl)$ is feasible, \ie
  there exists a hypergraph isomorphism $\phi$ such that $\cl_\phi =
  \cl$. Lemma~\ref{lem:noexit1} and Lemma~\ref{lem:xnotempty} show us
  that \filteri and \filterii do not exit if input is feasible. Thus
  Algorithm~\ref{al:icppl-find-isomorph} returns a non-empty
  function.%\qed
\end{proof}

@@

Now see
Figure~\ref{fig:solpartIII} to see how the algorithm proceeds.
Before that it can be observed in Figure~\ref{fig:solpartI}, that none of the
leaves are shared by more than one path.  Hence,
\filteri does not alter the hypergraphs. \filterii finds the leaf
assignments as shown -- elements \xPa, \xSn, \xLu, \xSc, \xLi, \xWo
are assigned to vertices 7, 8, 9, 11, 10, 4 respectively. Computation
of these leaf labels are deterministic except for $\xLu, \xSc$ where
the algorithm arbitrarily chooses them from $X = \{\xLu, \xSa, \xSc\}$
and for $\xSn, \xWo$ arbitrarily chooses from $X = \{\xSn,
\xWo\}$. Here we show only one such choice and clearly one could have
chosen another leaf label (\eg $\xSa$ on 9 or $\xWo$ on 8) to get a
different feasible hypergraph isomorphism.  After this, the leaves of
$T$ are pruned and the hypergraph updated accordingly by \algoiii
subroutine -- the pruned leaves are shown with dotted edges and the
corresponding Venn diagrams show the updated hypergraphs. $\phi_0$ is
this leaf assignment and the algorithm proceeds to solve the
subproblem with the new hypergraphs and path labeling which is:
$\{2,6,5\}, \{2\}, \{6,5,3\}, \{1,5,3\}$ labeled with $\{\xPi, \xVi,
\xCh\}, \{\xPi\}, \{\xCh, \xFr, \xVi\}, \{\xSa, \xCh, \xFr\}$
respectively.

@@

Section~\ref{sec:feasible} described properties that a TPL must have
for it to be feasible. The next problem of interest is to test if a
given hypergraph is a path hypergraph with respect to a given target
tree\footnote{A larger problem would be to find if a given hypergraph
  is a path hypergraph with respect to any tree. This problem is not
  addressed in this thesis.}.  In other words, the problem is to find
out if a feasible tree path labeling exists from a given target tree
for a given hypergraph. In this section we will see two special cases
of this problem where the target tree is from a particular family of
trees. The first one, where the tree is a path as described by the
\CFTPLINT problem (see Section~\ref{sec:myresearchintro}) is a shown
to be equivalent to the well studied problem of consecutive-ones in
Section~\ref{sec:icpplicpia}. The second one, where the tree is a
$k$-subdivided tree as described by \CFTPLKTREE (see
Section~\ref{sec:myresearchintro}), has been solved using a polynomial
time algorithm. The latter problem enforces some conditions on the
hypergraph too which will be seen in Section~\ref{sec:ksubdivstar}.

@@

In this section the given hypergraph, the $k$-subdivided star and the
root of the star are denoted by $\cO$, $T$ and vertex $r$,
respectively.
% For each hyperedge $X \in \cO$, we will maintain a 2-tuple of
% non-negative numbers $\seq{p_1(X), p_2(X)}$.  The numbers satisfy
% the property that $p_1(X) + p_2(X) \leq |X|$, and at the end of path
% labeling, for each $X$, $p_1(X) + p_2(X) = |X|$.  This signifies the
% algorithm tracking the lengths of subpaths of the path assigned to
% $X$ from at most two rays. We also maintain another parameter called
% the {\em residue} of $X$ denoted by $s(X)=|X| - p_1(X)$. This
% signifies the residue path length that must be assigned to $X$ which
% must be from another ray. For instance, if $X$ is labeled a path
% from only one ray, then $p_1(X) = |X|$, $p_2(X) = 0$ and $s(X) = 0$.

@@

The set $\cO_i$ refers to the set of hyperedges $\cT_1^i \cup \cT_2^i$
in the $i$th iteration.  Note that $\cO_1 = \cO$.  In the $i$th
iteration, hyperedges from $\cO_i$ are assigned paths from $R_i$ using
the following rules. Also the end of the iteration, $\cL_1^{i+1},
\cL_2^{i+1}, \cT_1^{i+1}, \cT_2^{i+1}$ are set to $\cL_1^{i},
\cL_2^{i}, \cT_1^{i}, \cT_2^{i}$ respectively, along with some
case-specific changes mentioned in the rules below.

>>>>

\begin{definition}[Path labeling of a hypergraph]
  \label{def:pathlabel}
  Consider a hypergraph $\cF$ % with universe $U$
  and a path system $\cP$ from tree $T$ such that $\bI(\cF) \cong
  \bI(\cP)$.  Then the associated bijection $\cl: \cF
  \rightarrow \cP$ due to this isomorphism is called a \textbf{path
    labeling} of $\cF$ from $T$. The path system $\cP$ is
  alternatively denoted as $\cF^\cl$. The path labeling may also be
  denoted by ordered pairs $(\cF,\cl)$ or $(\cF,\cl,T)$.
\end{definition}
% If the intersection graphs of $\cF$ and $\cP$ are isomorphic,
% $\bI(\cF) \cong \bI(\cP)$, then the associated bijection $\cl: \cF
% \rightarrow \cP$ due to this isomorphism is called a {\em path
%   labeling} of the hypergraph $\cF$.
To elaborate on the phrase ``associated bijection'' in
Definition~\ref{def:pathlabel}, let $\cg: V(\bI(\cF)) \rightarrow
V(\bI(\cP))$ be the given intersection graph isomorphism, where
$V(\bI(\cF)) = \{ v_S \mid S \in \cF\}$ and $V(\bI(\cP)) = \{v_P \mid
P \in \cP\}$. Then the associated bijection $\cl: \cF \rightarrow \cP$
is defined as follows. For all $S' \in \cF$ and $P' \in \cP$, $\cl(S')
= P'$ \iff $\cg(v_{S'}) = v_{P'}$.  Thus, it can be seen that a path
labeling preserves the non-empty intersection property between every
pair of hyperedges in $\cF$.

@@

\begin{definition}[Hypergraph isomorphism induced path labeling]
  \label{def:inducedpl}
  Let $\cF$ be a hypergraph and $\cP$ be a path system \stt $\cF \cong
  \cP$. Let  $\phi: \supt{\cF} \rightarrow \supt{\cP}$ be this hypergraph
  isomorphism function. % described in
  % Definition~\ref{def:pathhypergraph}.
  Then there is an \textbf{induced path labeling} % denoted by $\cl_\phi:
%   \cF \rightarrow \cP$ 
  denoted by $\cp_\phi: \cF \rightarrow \cP$
  of the hypergraph \stt % $\cl_\phi(S) = \phi(S)$
  $\cp_\phi(S) = \phi(S)$
  (see Definition~\ref{def:hisomorph}) for all $S \in \cF$.  
\end{definition}

@@

\begin{definition}[Feasible path labeling]
  \label{def:feasibletpl}
  Let $\cF$ be a hypergraph and $(\cF, \cl)$ be a path
  labeling. $(\cF, \cl)$ is called a \textbf{ feasible path labeling} if
  the following conditions hold.
  \begin{itemize}
  \item $\cF \cong \cF^\cl$
  \item if the above hypergraph isomorphism is $\phi: supp(\cF)
    \rightarrow supp(\cF^\cl)$, then $\cp_\phi = \cl$
  \end{itemize}

@@

Now we will see the definition of a \emph{ path hypergraph} which is a
simple idea extended from interval hypergraphs in \cite{kklv10}.

\begin{definition}[Path hypergraph]
  \label{def:pathhypergraph}
  Let $\cF$ be a set system. $\cF$ is called a \textbf{path hypergraph}
  if there exists a target tree $T$ and a path system $\cP$ from $T$
  \stt $\cF \cong \cP$. If $\cF$ is thus a path hypergraph, $\cP$ is
  called \textbf{path representation} of $\cF$.
\end{definition}

@@

\begin{definition}[$k$-subdivided star]
  \label{def:ksubstar}
  A \textbf{$k$-subdivided star} is a star with all its rays subdivided
  exactly $k$ times. A \textbf{ray} of a $k$-subdivided star is the path
  from the center vertex to a leaf. It is clear that all rays of a
  $k$-subdivided star are of length $k+2$.
\end{definition}

@@

\begin{definition}[In-tree]
An \textbf{in-tree} is a directed rooted tree in which all edges are
directed toward to the root.  
\end{definition}
In-trees are used in Section~\ref{sec:norestraint} which describes our
solution to \CFTPL problem.

@@

\begin{definition}
  \label{def:icppl}
  Consider a path labeling $(\cF, \cl)$ on the given tree $T$. We call
  $(\cF, \cl)$ an \textbf{Intersection Cardinality Preserving Path
    Labeling (ICPPL)} if it has the following properties.
  \par
  \parbox{\linewidth}{%
    \hangindent \defindent%
    \begin{enumerate}[i. ] %{(\icpplpr\ }
      \singlespacing
    \item \label{pr:i} $|S| = |\cl(S)|$ for all $S \in \cF$
    \item \label{pr:ii}$|S_1 \cap S_2| = |\cl(S_1) \cap \cl(S_2)|$ for
    all distinct $S_1, S_2 \in \cF$
    \item \label{pr:iii}$|S_1 \cap S_2 \cap S_3| = |\cl(S_1) \cap
    \cl(S_2) \cap \cl(S_3)|$ for all distinct $S_1, S_2, S_3 \in \cF$
    \end{enumerate}
  }
\end{definition}

@@

Finally we have Theorem~\ref{th:charac} which puts together all that
we saw so far in this section to prove that ICPPL is indeed a
characterization of a feasible path labeling.
\begin{theorem}
  \label{th:charac}
  A path labeling $(\cF, \cl)$ on tree $T$ is feasible \iff it is an
  ICPPL and Algorithm~\ref{al:icppl-find-isomorph} with $(\cF, \cl,
  T)$ as input returns a non-empty function.
\end{theorem}
\begin{proof}\thesisspacing
  From Lemma~\ref{lem:hyperiso}, we know that if $(\cF, \cl)$ is an
  ICPPL and if Algorithm~\ref{al:icppl-find-isomorph} with $(\cF, \cl,
  T)$ as input returns a non-empty function, then $(\cF, \cl)$ is
  feasible.  Now consider the case where $(\cF, \cl)$ is feasible, \ie
  there exists a hypergraph isomorphism $\phi$ such that $\cp_\phi =
  \cl$. 
  Algorithm~\ref{al:icppl-find-isomorph} then clearly returns a non-empty
  function because the target tree is non-empty.%\qed
\end{proof}

@@

Now see
Figure~\ref{fig:solpartIII} to see how the algorithm proceeds.
Before that it can be observed in Figure~\ref{fig:solpartI}, that none of the
leaves are shared by more than one path.  Hence,
\filteri does not alter the hypergraphs. \filterii finds the leaf
assignments as shown -- elements \xPa, \xSn, \xLu, \xSc, \xLi, \xWo
are assigned to vertices 7, 8, 9, 11, 10, 4 respectively. Computation
of these leaf labels are deterministic except for $\xLu, \xSc$ where
the algorithm arbitrarily chooses them from $X = \{\xLu, \xSa, \xSc\}$
and for $\xSn, \xWo$ arbitrarily choosen from $X = \{\xSn,
\xWo\}$. Here we show only one such choice and clearly one could have
chosen another leaf label (\eg $\xSa$ on 9 or $\xWo$ on 8) to get a
different feasible hypergraph isomorphism.  After this, the leaves of
$T$ are pruned and the hypergraph updated accordingly by \algoiii
subroutine -- the pruned leaves are shown with dotted edges and the
corresponding Venn diagrams show the updated hypergraphs. $\phi_0$ is
this leaf assignment and the algorithm proceeds to solve the
subproblem with the new hypergraphs and path labeling which is:
$\{2,6,5\}, \{2\}, \{6,5,3\}, \{1,5,3\}$ labeled with $\{\xPi, \xVi,
\xCh\}, \{\xPi\}, \{\xCh, \xFr, \xVi\}, \{\xSa, \xCh, \xFr\}$
respectively.

@@

Section~\ref{sec:feasible} described properties that a TPL must have
for it to be feasible. The next problem of interest is to test if a
given hypergraph is a path hypergraph with respect to a given target
tree\footnote{A larger problem would be to test if a given hypergraph
  is a path hypergraph where target tree is not given as input. This
  problem is not addressed in this thesis and is discussed as further
  research in Chapter~\ref{ch:conclusion}.}.  In other words, the
problem is to find out if a feasible tree path labeling exists from a
given target tree for a given hypergraph. In this section we will see
two special cases of this problem where the target tree is from a
particular family of trees. The first one, where the tree is a path as
described by the \CFTPLINT problem (see
Section~\ref{sec:myresearchintro} for definition) in
Section~\ref{sec:icpplicpia} is known to be equivalent to the well
studied problem of consecutive-ones. The second one, where the tree is
a $k$-subdivided tree as described by \CFTPLKTREE (see
Section~\ref{sec:myresearchintro} for definition), has been solved
using a polynomial time algorithm. The latter problem enforces some
conditions on the hypergraph too which will be seen in
Section~\ref{sec:ksubdivstar}.

@@

% For each hyperedge $X \in \cO$, we will maintain a 2-tuple of
% non-negative numbers $\seq{p_1(X), p_2(X)}$.  The numbers satisfy
% the property that $p_1(X) + p_2(X) \leq |X|$, and at the end of path
% labeling, for each $X$, $p_1(X) + p_2(X) = |X|$.  This signifies the
% algorithm tracking the lengths of subpaths of the path assigned to
% $X$ from at most two rays. We also maintain another parameter called
% the {\em residue} of $X$ denoted by $s(X)=|X| - p_1(X)$. This
% signifies the residue path length that must be assigned to $X$ which
% must be from another ray. For instance, if $X$ is labeled a path
% from only one ray, then $p_1(X) = |X|$, $p_2(X) = 0$ and $s(X) = 0$.

@@
